Surrogate model of aligned-spin binary neutron star waveforms using Gaussian process regression

### 2017-12-02 paper/paper.tex
Corrected TaylorF2 and verified with lalsimulation version.
Adding references.

For the parameter most sensitive to waveform errors, the tidal parameter, we find that the error from the resulting surrogate leads to a systematic error in the measured tidal parameter $\tilde\Lambda$ of less than $\sim 50$ which is less than 10\% of typical tidal parameters.

@@

The detection of the binary neutron star (BNS) system GW170817 [[cite]] by the Advanced LIGO [[cite]] and Advanced Virgo [[cite]] gravitational-wave detectors, has demonstrated that the sky position, distance, masses, spins, and tidal parameters can be measured with enough precision to answer fundamental physics questions such as identifying BNS mergers as a source of gamma ray bursts [[cite]], measuring the Hubble constant [[cite]], and understanding the structure of neutron stars (NSs) through tidal interactions [[cite]]. Measuring these parameters accurately will also be necessary to properly interpret the gamma ray burst [[cite]] and optical afterglow [[cite]] that was found in conjunction. Furthermore, an estimate of the BNS merger rate of x--y events Gpc$^{-3}$~yr$^{-1}$ (90\% confidence interval) [[GW170817]] indicates that, when second generation detectors reach design sensitivity, including KAGRA [[cite]] and possibly LIGO-India [[cite]], we could eventually observe a population of tens or hundreds of events after a couple of years of observation [[2010Rates?]]. This will allow us to estimate the NS mass distribution [[cite]], spin distribution [[cite]], and universal equation of state (EOS) that is common to all NSs [[DelPozzo, Lackey, ...]].

@@

The effective one body (EOB) formalism, on the other hand, has significantly smaller uncertainties than PN waveforms. Variations within the EOB formalism lead to uncertainties in the waveform phase of $\sim 1$~rad [[cite or say as shown below]] as compared to PN waveform uncertainties of $\sim 10$~rad. Furthermore, NR simulations of BNS systems now claim phase errors of less than 1~rad during the last $\sim 10$ orbits through merger [[cite]] and show good agreement with EOB models [[cite]].

@@

Surrogate modeling techniques have had significant success in gravitational-wave data analysis for accelerating the evaluation of a waveform $h(t;\bx)$ given its waveform parameters $\bx$. The standard technique is to construct a training set of waveforms that sample the parameter space, then compress this training set by building a reduced, orthonormal basis whose linear combination can reproduce any waveform in the parameter space to some desired accuracy. This serves to interpolate the waveform as a function of time (or frequency for frequency-domain waveforms) in terms of a reduced set of coefficients corresponding to the reduced set of basis functions. In previous work, this reduced basis has been constructed either with singular value decomposition (SVD) [[Michael's papers]] or with a greedy method that chooses a subset of the training-set waveforms that are as orthogonal to each other as possible, then orthonormalizes this subset with Graham-Schmidt orthogonalization [[All the other papers]].

@@

Previous work on surrogate models has focused mainly on parameter spaces $\bx$ with $\le 3$~dimensions. This allows one to approximate the coefficients $c_i(\bx)$ or waveform values $h(T_j; \bx)$ as functions of $\bx$ using simple fitting functions [[cite]] or standard interpolation techniques such as tensor spline [[Michael, others]] or Chebyshev interpolation [[Ben]]. These interpolation techniques typically require waveforms evaluated on a rectangular grid, and this suffers from the curse of dimensionality; for $N$ points per dimension $d$,  $N^d$ waveforms are needed. For the 5-dimensional problem here, this is somewhat unfeasable. Recent work on optimally choosing waveforms for both EOB [[cite GPRpaper]] and NR [[cite CaltechPrecessing]] waveforms has shown that there are sufficiently accurate alternatives to grid-based interpolation that do not suffer from the curse of dimensionality.

In this paper we focus on breaking this curse of dimensionality by using an interpolation method known as Gaussian process regression (GPR) that does not require a regular grid. Importantly, GPR also provides a convenient estimate of its own uncertainty. This allows us to iteratively add new points to the training set in such a way as to minimize the interpolation error over the parameter space for a given number of waveforms in the training set. A similar approach has recently been used to optimally sample waveforms for a 2-dimensional parameter space using GPR [[DoctorEtAl]]. GPR has also been used recently in GW data analysis to marginalize over waveform uncertainties in parameter estimation [[ChrisMoore2papers?]].

In this paper, we will develop methods to construct a surrogate model for the inspiral through merger of an aligned spin BNS waveform with tidal interactions using the EOB tidal description described in Hinderer et al.~[[cite1, cite2]]. This work is a continuation of the surrogate developed in [[cite]] for a nonspinning BNS waveform using a slightly different description for the point-particle and tidal pieces by Bernuzzi et al.~[[cite, ...]]. Whereas in [[Ben]] we constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model which significantly speeds up parameter estimation by not requiring an online Fourier transform for each waveform evaluation. It also enables additional techniques such as reduced order quadrature [[cite]] to accelerate likelihood evaluation.

@@

%Because the waveform $\tilde h(Mf; \bx)$ is an oscillatory function of $Mf$ and $\bx$, the waveform is usually decomposed into an amplitude $A(Mf; \bx)$ and phase $\Phi(Mf; \bx)$ as $\tilde h(Mf; \bx) = A(Mf; \bx) e^{i\Phi(Mf; \bx)}$. These functions are smoother, mostly monotonic functions of $Mf$ and $\bx$. This can be seen in Fig.~\ref{fig:h} which shows the initial training set of waveforms. \red{[[Actually, you might want to show the 51 waveforms with only $S_1$ varied. Then show them as a function of $f$ and another plot as a function of $S_1$.]]} Unfortunately, the amplitude and phase still span a wide range of values. The phase, for example, spans about $10^4$~rad. To avoid systematic errors in the tidal parameters, we need phase errors of $\lesssim 1$~rad over most of this frequency range, leading to a requirement on the fractional interpolation error of $\lesssim 10^{-4}$~rad. This is a difficult requirement to achieve in the 5-dimensional parameter space here.

%\begin{figure}[htb]
%\centering
%\includegraphics[width=0.49\textwidth]{h_trainingset.png}
%%\includegraphics[width=0.49\textwidth]{dh_trainingset.png}
%\caption{Training set. Maybe instead show the waveform set which varies 1 parameter.}
%\label{fig:h}
%\end{figure}

%\begin{figure}[htb]
%\centering
%%\includegraphics[width=0.49\textwidth]{h_trainingset.png}\\
%\includegraphics[width=0.49\textwidth]{dh_trainingset.png}
%\caption{Training set. Maybe instead show the waveform set which varies 1 parameter.}
%\label{fig:dh}
%\end{figure}

%To demonstrate the usefulness of this approach, in Fig.~\ref{fig:phaserange}, we show the range of values of the phase $\Phi$ over the training set $\Phi_{\rm range} = \Phi_{\rm max} - \Phi_{\rm min}$ as well as the range of values of the phase residual $\Delta\Phi$ over the training set $\Delta\Phi_{\rm range} = \Delta\Phi_{\rm max}-\Delta\Phi_{\rm min}$. The range in $\Phi$ can be larger than 1000~radians. The range of $\Delta\Phi$ is 1--2 orders of magnitude smaller, and as a result, the interpolation accuracy requirements are 1--2 orders of magnitude smaller as well. 
%
%\begin{figure}[htb]
%\centering
%\includegraphics[width=0.49\textwidth]{phaserange.png}
%\caption{Range of phase values for training set waveforms. The range is a factor of $\gtrsim 100$ larger for $\Phi$ than for $\Delta\Phi$, 
%meaning that the interpolation accuracy requirements are at least $\sim 2$ orders of magnitude less for $\Delta\Phi$.}
%\label{fig:phaserange}
%\end{figure}

@@

For aligned-spin waveforms, we can solve this problem using the fact that the waveform is approximately known analytically from the stationary phase approximation to the post-Newtonian (PN) waveform known as TaylorF2 [[cite]] $\tilde h_{\rm F2}(Mf; \bx) = A_{\rm F2}(Mf; \bx) e^{i\Phi_{\rm F2} (Mf; \bx)}$. This allows us to write the EOB amplitude and phase in terms of small residuals, $\Delta\ln(A)(Mf; \bx)$ and $\Delta\Phi(Mf; \bx)$, relative to TaylorF2
\begin{align}
\Delta\ln(A)(Mf; \bx) &= \ln\left(\frac{ A(Mf; \bx) }{ A_{\rm F2}(Mf; \bx)}\right), \label{eq:ampres}\\
\Delta\Phi(Mf; \bx) & = \Phi(Mf; \bx) - \Phi_{\rm F2}(Mf; \bx), \label{eq:phaseres}
\end{align}
such that
\begin{align}
\tilde h(Mf; \bx) = \tilde h_{\rm F2}(Mf; \bx) e^{\Delta\ln(A)(Mf; \bx) + i  \Delta\Phi(Mf; \bx)}.
\label{eq:hdecomp}
\end{align}
These residuals are also shown in Fig.~\ref{fig:hoff}. For the amplitude residual, we use a log-ratio instead of a ratio because it guarantees that interpolation errors will not lead to a negative amplitude for the reconstructed waveform (Eq.~\eqref{eq:hdecomp}). In addition, because the waveform amplitude spans several orders of magnitude after the merger frequency, the log-ratio captures this behavior. Comparing the phase $\Phi$ and phase residual $\Delta\Phi$ in Fig.~\ref{fig:hoff}, we find that the range in $\Delta\Phi$ is a few orders of magnitude smaller than the range in $\Phi$ except at very high frequencies. In Fig.~\ref{fig:dhofs}, we show the amplitude and phase residuals at three fixed frequencies as functions of the waveform parameter $S_1$. These quantities are smooth functions of $\bx$, implying that interpolation as a function of $\bx$ should work. The exception is the amplitude residual at very high frequencies where the amplitude is small and numerical noise dominates.

@@

The explicit expressions for the amplitude and phase of the TaylorF2 waveform are as follows. For the amplitude we use the 1PN correction to the leading order waveform.
\begin{equation}
A_{\rm F2} = \sqrt{\frac{5\pi\eta}{24}} x^{-7/4} \left[ 1 + \left(-\frac{323}{224} + \frac{451\eta}{168} \right)x \right],
\end{equation}
where $x=(\pi M f)^{2/3}$ is the standard PN parameter. This differs slightly from the decomposition used by \texttt{lalsuite}. The phase has the schematic form
\begin{align}
\Phi_{\rm F2} =& -2\pi f t_c + \phi_c + \frac{\pi}{4} - \frac{3}{128\eta}x^{-5/2} \left[1 + \right. \\
                        & \left. {\rm PP}(\eta) + {\rm Spin}(\eta, S_1, S_2) + {\rm Tidal}(\eta, \Lambda_1, \Lambda_2)  \right].
\end{align}
We use the point particle terms ${\rm PP}(\eta)$ to 3.5PN order in Eq.~[[]] of Ref.~[[]], the aligned spin terms ${\rm Spin}(\eta, S_1, S_2)$ to 3PN order in Eqs.~[[]] of Ref.~[[]] and Eqs.~[[]] of Ref.~[[]], and the tidal terms ${\rm Tidal}(\eta, \Lambda_1, \Lambda_2)$ to 6PN order in Eqs.~[[]] of Ref.~[[]]. These terms for the phase are exactly as used in the \texttt{lalsuite} waveform \texttt{TaylorF2} [[cite]].

@@

We evaluate the EOB waveform with a starting frequency of $Mf_{{\rm win}, i}=0.000197$, equivalent to a physical frequency of 20~Hz for a binary with total mass $M=2M_\odot$. In order to take a discrete Fourier transform, we window the start of the waveform with a Planck window [[cite]] in the interval $[Mf_{{\rm win}, i}, Mf_{{\rm win}, f}] = [0.000197, 0.00021]$ to avoid Gibbs oscillations. The end of the waveform has zero amplitude, so the end does not need to be windowed. We then resample the waveform with a spacing $\Delta t/M = 5$ and pad the end of the waveform with zeros such that all waveforms in the training set have the exact same time samples. After evaluating the discrete Fourier transform, we calculate the residuals $\Delta\ln(A)(Mf)$ and $\Delta\Phi(Mf)$ between the EOB and TaylorF2 waveforms using Eqs.~\eqref{eq:ampres} and~\eqref{eq:phaseres}.

@@

We now seek to interpolate the conditioned residuals $\Delta\ln(A)(Mf; \bx)$ and $\Delta\Phi(Mf; \bx)$. We begin by choosing a method for interpolating as a function of $Mf$ for fixed $\bx$. The majority of previous papers on gravitational-wave surrogates have used an orthonormal basis of global functions $\hat e_i(Mf)$ for interpolating as a function of frequency (or time) [[cite]]. This type of surrogate, built from a reduced basis, is usually referred to as a reduced order model. For a generic function $g(Mf; \bx)$, this decomposition can be written as $g(Mf; \bx) \approx \sum_{i=1}^N c_i(\bx) \hat e_i(Mf)$. The coefficients $c_i(\bx)$ are then interpolated as functions of $\bx$ [[Puerrer]]. Alternatively, using the empirical interpolation method [[cite]], one can re-express these basis functions $\hat e_i(Mf)$ in terms of empirical interpolating functions $B_j(Mf)$ and the value of $g(Mf; \bx)$ at empirical nodes $MF_j$ as $g(Mf; \bx) \approx \sum_{j=1}^N B_j(Mf) g(MF_j; \bx)$. The location of these nodes $MF_j$ is then optimized to minimize interpolation errors. This method has been used in Refs.~[[cite]].

@@

\red{[[Make sure all vectors and matrices are bold.]]}

Next we choose a method to interpolate the residuals $\Delta\ln(A)(MF_j; \bx)$ and $\Delta\Phi(MF_j; \bx)$ at each of the frequency nodes $MF_j$ as a function of the five waveform parameters $\bx$. Most multivariate interpolation techniques (e.g. tensor spline or Chebyshev interpolation) require a function to be sampled on a rectangular grid, and thus suffer from the curse of dimensionality: the number of samples grows exponentially with the dimension $d$ ($N^d$ samples for $N$ points per dimension). For our 5-dimensional problem, $10^5$ waveform evaluations are needed for only 10 samples per parameter. At $\sim10$~minutes--1~day per EOB waveform on a standard CPU, this is at the limit of what is reasonable. However, there are other methods that do not require a rectangular grid, and this allows us to choose more efficient experimental designs as discussed in Sec.~\ref{sec:design} below. The method we choose is Gaussian Process Regression (GPR).

In GPR a function $g(\bx)$ is described in terms of a mean $m(\bx)$ and  covariance $k(\bx, \bx')$ between points [[cite RassmussenWilliamsChapter2, IntroArxivPaper]].
\begin{equation}
g(\bx) \approx \mathcal{GP}(m(\bx), k(\bx, \bx')).
\end{equation}
The mean $m(\bx)$ can be a parameterized fitting function while $k(\bx, \bx')$ is a kernel with tunable hyperparameters that describes the covariance between the point $\bx$ and the sampled point $\bx'$. This covariance can be used to describe features of the function not captured by the parameterized mean $m(\bx)$ as well as genuine noise in the data $\by$. For the problem here, we have already subtracted the TaylorF2 waveform from EOB waveform, so we set $m(\bx) = 0$ and model the residuals purely in terms of the kernel $k(\bx, \bx')$.

@@

We use a radial kernel which expresses the covariance in terms of a distance $r$ between points
\begin{equation}
r^2 = (\bx - \bx')^T M (\bx - \bx'),
\end{equation}
where we choose the matrix $M$ to be diagonal
\begin{equation}
M = {\rm diag}(\ell_1^{-2}, \ell_2^{-2}, \dots, \ell_d^{-2}).
\end{equation}
The tunable hyperparameters $\ell_i$ represent the length scale over which the function $g(\bx)$ varies in each coordinate $x_i$. Specifically, we use the Mat\'{e}rn class of kernels
\begin{equation}
k_{\rm Matern}(r) = \frac{2^{1-\nu}}{\Gamma(\nu)} \left(\sqrt{2\nu} r\right)^\nu K_\nu \left(\sqrt{2\nu} r \right).
\end{equation}
where $K_\nu(x)$ is a modified Bessel function. The value of $\nu$ parameterizes the smoothness of the Gaussian process, and is $k$ times mean-square differentiable if $\nu>k$ [[cite]]. For half-integer values of $\nu$, this kernel has a computationally cheap form without special functions, and we have had good results with $\nu=5/2$, resulting in a twice-differentiable function\footnote{If desired, one can select the optimal class of kernels using cross validation, and we have also tried the more common, infinitely differentiable squared exponential kernel [[cite]]. However, we found it difficult to reliably tune its hyperparameters for the problem here.}. The $\nu=5/2$ kernel is
\begin{equation}
k_{\rm Matern}^{\nu=5/2}(r) = \left(1+\sqrt{5}r + \frac{5r^2}{3}\right) \exp\left(-\sqrt{5}r\right).
\end{equation}

@@

In order to estimate the hyperparameters, we use the above assumption (Eq.~\eqref{eq:gaussian}) that the joint distribution of 
the data ${\bm y}$ is a multivariate Gaussian which has the following distribution
\begin{equation}
\ln p({\bm y} | {\bm x}, {\bm \theta}) = -\frac{1}{2}{\bm y}^T K^{-1} {\bm y} - \frac{1}{2} \ln |K| - \frac{d}{2} \ln 2\pi.
\end{equation}
This is the log-likelihood for ${\bm y}$ given the hyperparameters ${\bm \theta}$, and we can find the posterior for ${\bm \theta}$
given ${\bm y}$ using Bayes' theorem
\begin{equation}
p({\bm \theta} | {\bm x}, {\bm y}) \propto p({\bm \theta}) p({\bm y} | {\bm x}, {\bm \theta}).
\end{equation}
The prior $p({\bm \theta})$ is typically uniform and used to set the bounds on ${\bm \theta}$. One can sample this posterior if interested in the distribution of hyperparameters $\btheta$. However, for the problem here, we simply want the maximum posterior. We do this using the \texttt{gaussian\_process} module in the \texttt{scikit-learn} package [[cite]]. With these optimized hyperparameters, the final interpolating function is given by Eq.~\eqref{eq:mean} and its uncertainty by Eq.~\eqref{eq:var}. We note that once the vector $K^{-1}y$ is precomputed, evaluating Eq.~\eqref{eq:mean} is a fast $\mathcal{O}(N)$ operation.

@@

%One of the main aims of this paper is to build a surrogate with as few waveform evaluations as possible. EOB models are expensive, and we would like a method that extends to higher dimensions and more expensive waveform models such as from numerical relativity simulations. Using the freedom provided by GPR to sample waveforms at arbitrary locations in parameter space, we try a set of more optimal designs.

>>>>

%For the parameter most sensitive to waveform errors, the tidal parameter, we find that the error from the resulting surrogate leads to a systematic error in the measured tidal parameter $\tilde\Lambda$ of less than $\sim 50$ which is less than 10\% of typical tidal parameters.

@@

The detection of the binary neutron star (BNS) system GW170817~\cite{GW170817} by the Advanced LIGO~\cite{Harry2010} and Advanced Virgo~\cite{Acernese2009} gravitational-wave detectors, has demonstrated that the sky position, distance, masses, spins, and tidal parameters can be measured with enough precision to answer fundamental physics questions such as identifying BNS mergers as a source of gamma ray bursts [[cite]], measuring the Hubble constant~\cite{GW170817Hubble}, and understanding the structure of neutron stars (NSs) through tidal interactions~\cite{GW170817}. Measuring these parameters accurately will also be necessary to properly interpret the gamma ray burst [[cite]] and optical afterglow [[cite]] that was found in conjunction. Furthermore, an estimate of the BNS merger rate of 320--4740 events Gpc$^{-3}$~yr$^{-1}$ (90\% confidence interval)~\cite{GW170817} indicates that, when second generation detectors reach design sensitivity, including KAGRA~\cite{Somiya2012} and possibly LIGO-India~\cite{IyerSouradeepUnnikrishnan2011}, we could eventually observe a population of tens or hundreds of events after a couple of years of observation~\cite{LIGORate2010}. This will allow us to estimate the NS mass distribution [[cite]], spin distribution [[cite]], and universal equation of state (EOS) that is common to all NSs\cite{DelPozzoLiAgathos2013, LackeyWade2015}.

@@

The effective one body (EOB) formalism, on the other hand, has significantly smaller uncertainties than PN waveforms. Variations within the EOB formalism lead to uncertainties in the waveform phase of $\sim 1$~rad [[cite or say as shown below]] as compared to PN waveform uncertainties of $\sim 10$~rad~\cite{HindererLackeyLangRead2010}. Furthermore, NR simulations of BNS systems now claim phase errors of less than 1~rad during the last $\sim 10$ orbits through merger [[cite]] and show good agreement with EOB models [[cite]].

@@

Surrogate modeling techniques have had significant success in gravitational-wave data analysis for accelerating the evaluation of a waveform $h(t;\bx)$ given its waveform parameters $\bx$. The standard technique is to construct a training set of waveforms that sample the parameter space, then compress this training set by building a reduced, orthonormal basis whose linear combination can reproduce any waveform in the parameter space to some desired accuracy. This serves to interpolate the waveform as a function of time (or frequency for frequency-domain waveforms) in terms of a reduced set of coefficients corresponding to the reduced set of basis functions. In previous work, this reduced basis has been constructed either with singular value decomposition (SVD)~\cite{Puerrer2014, Puerrer2015} or with a greedy method that chooses a subset of the training-set waveforms that are as orthogonal to each other as possible, then orthonormalizes this subset with Graham-Schmidt orthogonalization~\cite{FieldGalleyHesthaven2014, LackeyBernuzziGalley2017, BlackmanFieldGalley2015, BlackmanFieldScheel2017a, BlackmanFieldScheel2017b}.

@@

Previous work on surrogate models has focused mainly on parameter spaces $\bx$ with $\le 3$~dimensions. This allows one to approximate the coefficients $c_i(\bx)$ or waveform values $h(T_j; \bx)$ as functions of $\bx$ using simple fitting functions [[cite]] or standard interpolation techniques such as tensor spline~\cite{Puerrer2014, Puerrer2015} or Chebyshev interpolation~\cite{LackeyBernuzziGalley2017}. These interpolation techniques typically require waveforms evaluated on a rectangular grid, and this suffers from the curse of dimensionality; for $N$ points per dimension $d$,  $N^d$ waveforms are needed. For the 5-dimensional problem here, this is somewhat unfeasable. Recent work on optimally choosing waveforms for both EOB~\cite{DoctorFarrHolz2017} and NR~\cite{BlackmanFieldScheel2017a, BlackmanFieldScheel2017b} waveforms has shown that there are sufficiently accurate alternatives to grid-based interpolation that do not suffer from the curse of dimensionality.

In this paper we focus on breaking this curse of dimensionality by using an interpolation method known as Gaussian process regression (GPR)~\cite{RasmussenWilliams2006} that does not require a regular grid. Importantly, GPR also provides a convenient estimate of its own uncertainty. This allows us to iteratively add new points to the training set in such a way as to minimize the interpolation error over the parameter space for a given number of waveforms in the training set. Similar approaches have recently been used to optimally sample waveforms for a 2-dimensional aligned spin BBH surrogate using GPR~\cite{DoctorFarrHolz2017} as well as for a surrogate of eccentric BBH mergers~\cite{HuertaMooreKumar2017}. GPR has also been used recently in GW data analysis to marginalize over waveform uncertainties in parameter estimation~\cite{MooreGair2014, MooreBerryChua2016}.

In this paper, we will develop methods to construct a surrogate model for the inspiral through merger of an aligned spin BNS waveform with tidal interactions using the EOB tidal description described in Hinderer et al.~[[cite1, cite2]]. This work is a continuation of the surrogate developed in Ref.~\cite{LackeyBernuzziGalley2017} for a nonspinning BNS waveform using a slightly different description for the point-particle and tidal pieces by Bernuzzi et al.~\cite{Bernuzzi:2014owa}. Whereas in Ref.~\cite{LackeyBernuzziGalley2017} we constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model which significantly speeds up parameter estimation by not requiring an online Fourier transform for each waveform evaluation. It also enables additional techniques such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015} to accelerate likelihood evaluation.

@@

@@

For aligned-spin waveforms, we can solve this problem using the fact that the waveform is approximately known analytically from the stationary phase approximation to the post-Newtonian (PN) waveform known as TaylorF2~\cite{BuonannoIyerOchsner2009} $\tilde h_{\rm F2}(Mf; \bx) = A_{\rm F2}(Mf; \bx) e^{i\Phi_{\rm F2} (Mf; \bx)}$. This allows us to write the EOB amplitude and phase in terms of small residuals, $\Delta\ln(A)(Mf; \bx)$ and $\Delta\Phi(Mf; \bx)$, relative to TaylorF2
\begin{align}
\Delta\ln(A)(Mf; \bx) &= \ln\left(\frac{ A(Mf; \bx) }{ A_{\rm F2}(Mf; \bx)}\right), \label{eq:ampres}\\
\Delta\Phi(Mf; \bx) & = \Phi(Mf; \bx) - \Phi_{\rm F2}(Mf; \bx), \label{eq:phaseres}
\end{align}
such that
\begin{align}
\tilde h(Mf; \bx) = \tilde h_{\rm F2}(Mf; \bx) e^{\Delta\ln(A)(Mf; \bx) + i  \Delta\Phi(Mf; \bx)}.
\label{eq:hdecomp}
\end{align}
These residuals are also shown in Fig.~\ref{fig:hoff}. For the amplitude residual, we use a log-ratio instead of a ratio because it guarantees that interpolation errors will not lead to a negative amplitude for the reconstructed waveform (Eq.~\eqref{eq:hdecomp}). In addition, because the waveform amplitude spans several orders of magnitude after the merger frequency, the log-ratio captures this behavior. Comparing the phase $\Phi$ and phase residual $\Delta\Phi$ in Fig.~\ref{fig:hoff}, we find that the range in $\Delta\Phi$ is a few orders of magnitude smaller than the range in $\Phi$ except at very high frequencies. In Fig.~\ref{fig:dhofs}, we show the amplitude and phase residuals at three fixed frequencies as functions of the waveform parameter $S_1$. These quantities are smooth functions of $\bx$, implying that interpolation as a function of $\bx$ should work. The exception is the amplitude residual at very high frequencies where the amplitude is small and numerical noise dominates.

@@

The explicit expressions for the amplitude and phase of the TaylorF2 waveform are as follows. For the amplitude we use the 1PN correction to the leading order waveform.
\begin{equation}
A_{\rm F2} = \sqrt{\frac{5\pi\eta}{24}} x^{-7/4} \left[ 1 + \left(-\frac{323}{224} + \frac{451\eta}{168} \right)x \right],
\end{equation}
where $x=(\pi M f)^{2/3}$ is the standard PN parameter. This differs slightly from the decomposition used by \texttt{lalsuite}~\cite{lal}. The phase has the schematic form
\begin{align}
\Phi_{\rm F2} =& -2\pi f t_c + \phi_c + \frac{\pi}{4} - \frac{3}{128\eta}x^{-5/2} \left[1 + \right. \\
                        & \left. {\rm PP}(\eta) + {\rm Spin}(\eta, S_1, S_2) + {\rm Tidal}(\eta, \Lambda_1, \Lambda_2)  \right].
\end{align}
We use the point particle terms ${\rm PP}(\eta)$ to 3.5PN order in Eq.~(3.18) of Ref.~\cite{BuonannoIyerOchsner2009}, the aligned spin terms ${\rm Spin}(\eta, S_1, S_2)$ to 3PN order~\cite{BoheMarsatBlanchet2013}, and the tidal terms ${\rm Tidal}(\eta, \Lambda_1, \Lambda_2)$ to 6PN order~\cite{VinesFlanaganHinderer2011}. These terms for the phase are exactly as used in the \texttt{lalsuite} waveform \texttt{TaylorF2}.

@@

We evaluate the EOB waveform with a starting frequency of $Mf_{{\rm win}, i}=0.000197$, equivalent to a physical frequency of 20~Hz for a binary with total mass $M=2M_\odot$. In order to take a discrete Fourier transform, we window the start of the waveform with a Planck window~\cite{McKechanRobinsonSathyaprakash2010} in the interval $[Mf_{{\rm win}, i}, Mf_{{\rm win}, f}] = [0.000197, 0.00021]$ to avoid Gibbs oscillations. The end of the waveform has zero amplitude, so the end does not need to be windowed. We then resample the waveform with a spacing $\Delta t/M = 5$ and pad the end of the waveform with zeros such that all waveforms in the training set have the exact same time samples. After evaluating the discrete Fourier transform, we calculate the residuals $\Delta\ln(A)(Mf)$ and $\Delta\Phi(Mf)$ between the EOB and TaylorF2 waveforms using Eqs.~\eqref{eq:ampres} and~\eqref{eq:phaseres}.

@@

We now seek to interpolate the conditioned residuals $\Delta\ln(A)(Mf; \bx)$ and $\Delta\Phi(Mf; \bx)$. We begin by choosing a method for interpolating as a function of $Mf$ for fixed $\bx$. The majority of previous papers on gravitational-wave surrogates have used an orthonormal basis of global functions $\hat e_i(Mf)$ for interpolating as a function of frequency (or time)~\cite{Puerrer2014, Puerrer2015, FieldGalleyHesthaven2014, LackeyBernuzziGalley2017, BlackmanFieldGalley2015, BlackmanFieldScheel2017a, BlackmanFieldScheel2017b}. This type of surrogate, built from a reduced basis, is usually referred to as a reduced order model. For a generic function $g(Mf; \bx)$, this decomposition can be written as $g(Mf; \bx) \approx \sum_{i=1}^N c_i(\bx) \hat e_i(Mf)$. The coefficients $c_i(\bx)$ are then interpolated as functions of $\bx$~\cite{Puerrer2014, Puerrer2015}. Alternatively, using the empirical interpolation method~\cite{Barrault2004, Chaturantabut2010, FieldGalleyHesthaven2014}, one can re-express these basis functions $\hat e_i(Mf)$ in terms of empirical interpolating functions $B_j(Mf)$ and the value of $g(Mf; \bx)$ at empirical nodes $MF_j$ as $g(Mf; \bx) \approx \sum_{j=1}^N B_j(Mf) g(MF_j; \bx)$. The location of these nodes $MF_j$ is then optimized to minimize interpolation errors.

@@

Next we choose a method to interpolate the residuals $\Delta\ln(A)(MF_j; \bx)$ and $\Delta\Phi(MF_j; \bx)$ at each of the frequency nodes $MF_j$ as a function of the five waveform parameters $\bx$. Most multivariate interpolation techniques (e.g. tensor spline or Chebyshev interpolation) require a function to be sampled on a rectangular grid, and thus suffer from the curse of dimensionality: the number of samples grows exponentially with the dimension $d$ ($N^d$ samples for $N$ points per dimension). For our 5-dimensional problem, $10^5$ waveform evaluations are needed for only 10 samples per parameter. At $\sim10$~minutes--1~day per EOB waveform on a standard CPU, this is at the limit of what is reasonable. However, there are other methods that do not require a rectangular grid, and this allows us to choose more efficient experimental designs as discussed in Sec.~\ref{sec:design} below. The method we choose is Gaussian Process Regression (GPR)~\cite{RasmussenWilliams2006}.

In GPR a function $g(\bx)$ is described in terms of a mean $m(\bx)$ and  covariance $k(\bx, \bx')$ between points.
\begin{equation}
g(\bx) \approx \mathcal{GP}(m(\bx), k(\bx, \bx')).
\end{equation}
The mean $m(\bx)$ can be a parameterized fitting function while $k(\bx, \bx')$ is a kernel with tunable hyperparameters that describes the covariance between the point $\bx$ and the sampled point $\bx'$. This covariance can be used to describe features of the function not captured by the parameterized mean $m(\bx)$ as well as genuine noise in the data $\by$. For the problem here, we have already subtracted the TaylorF2 waveform from EOB waveform, so we set $m(\bx) = 0$ and model the residuals purely in terms of the kernel $k(\bx, \bx')$.

@@

We use a radial kernel which expresses the covariance in terms of a distance $r$ between points
\begin{equation}
r^2 = (\bx - \bx')^T M (\bx - \bx'),
\end{equation}
where we choose the matrix $M$ to be diagonal
\begin{equation}
M = {\rm diag}(\ell_1^{-2}, \ell_2^{-2}, \dots, \ell_d^{-2}).
\end{equation}
The tunable hyperparameters $\ell_i$ represent the length scale over which the function $g(\bx)$ varies in each coordinate $x_i$. Specifically, we use the Mat\'{e}rn class of kernels
\begin{equation}
k_{\rm Matern}(r) = \frac{2^{1-\nu}}{\Gamma(\nu)} \left(\sqrt{2\nu} r\right)^\nu K_\nu \left(\sqrt{2\nu} r \right).
\end{equation}
where $K_\nu(x)$ is a modified Bessel function. The value of $\nu$ parameterizes the smoothness of the Gaussian process, and is $k$ times mean-square differentiable if $\nu>k$~\cite{RasmussenWilliams2006}. For half-integer values of $\nu$, this kernel has a computationally cheap form without special functions, and we have had good results with $\nu=5/2$, resulting in a twice-differentiable function\footnote{If desired, one can select the optimal class of kernels using cross validation~\cite{RasmussenWilliams2006}, and we have also tried the more common, infinitely differentiable squared exponential kernel~\cite{RasmussenWilliams2006}. However, we found it difficult to reliably tune its hyperparameters for the problem here.}. The $\nu=5/2$ kernel is
\begin{equation}
k_{\rm Matern}^{\nu=5/2}(r) = \left(1+\sqrt{5}r + \frac{5r^2}{3}\right) \exp\left(-\sqrt{5}r\right).
\end{equation}

@@

In order to estimate the hyperparameters, we use the above assumption (Eq.~\eqref{eq:gaussian}) that the joint distribution of 
the data ${\bm y}$ is a multivariate Gaussian which has the following distribution
\begin{equation}
\ln p({\bm y} | {\bm x}, {\bm \theta}) = -\frac{1}{2}{\bm y}^T K^{-1} {\bm y} - \frac{1}{2} \ln |K| - \frac{d}{2} \ln 2\pi.
\end{equation}
This is the log-likelihood for ${\bm y}$ given the hyperparameters ${\bm \theta}$, and we can find the posterior for ${\bm \theta}$
given ${\bm y}$ using Bayes' theorem
\begin{equation}
p({\bm \theta} | {\bm x}, {\bm y}) \propto p({\bm \theta}) p({\bm y} | {\bm x}, {\bm \theta}).
\end{equation}
The prior $p({\bm \theta})$ is typically uniform and used to set the bounds on ${\bm \theta}$. One can sample this posterior if interested in the distribution of hyperparameters $\btheta$. However, for the problem here, we simply want the maximum posterior. We do this using the \texttt{gaussian\_process} module in the \texttt{scikit-learn} package~\cite{scikit-learn}. With these optimized hyperparameters, the final interpolating function is given by Eq.~\eqref{eq:mean} and its uncertainty by Eq.~\eqref{eq:var}. We note that once the vector $K^{-1}y$ is precomputed, evaluating Eq.~\eqref{eq:mean} is a fast $\mathcal{O}(N)$ operation.

@@

One of the main aims of this paper is to build a surrogate with as few waveform evaluations as possible. Using the freedom provided by GPR to sample waveforms at arbitrary locations, we try a set of designs more efficient than uniform grids. One such design is a Latin Hypercube Design (LHD)~\cite{McKayBeckmanConover1979}. An LHD with $N$ samples divides each of $d$ dimensions uniformly into $N$ grid points for a total of $N^d$ grid points. However, unlike a uniform grid, the $N$ values in each dimension are sampled exactly once instead of $N^{d-1}$ times. For an LHD there are $(N!)^d$ ways to choose these points, and we choose one randomly\footnote{There are ways to better choose an LHD. A standard requirement is that the LHD be space-filling, meaning that the points are as far apart as possible from each other. (See Ref.~\cite{Husslage2011} for a review of methods for optimizing the placement of samples.) One such definition of space filling is that the chosen locations maximize the minimum Euclidean distance between any two samples. We have not experimented with these alternatives here.}.
